---
title: "Ch 6: Macro"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

+ You can summarize the state of the macroeconomy with just a few numbers:
    + Inflation -- How all prices are changing
    + GDP Growth -- How production of goods and services is changing
    + Unemployment -- People that are looking for work.

+ Each one has a lot of details, definitions and alternative measures
+ How we think of them has changed over time
+ All are generated by surveys.  There is no dial on the economy.

## Today

```{r message=FALSE, warning=FALSE, include=FALSE}
library(alfred)
library(dplyr)
library(ggplot2)

LatestUnemployment <- last(get_alfred_series("UNRATE")$UNRATE)
LatestGDPGrowth <- get_alfred_series("A191RL1Q225SBEA") %>% last() %>% last()
LatestInflation <- get_alfred_series("CPALTT01USM659N") %>% last() %>% last()

```
+ Unemployment Rate Last Month: `r LatestUnemployment`\%
+ GDP Growth (From Previous Quarter Annualized): `r LatestGDPGrowth`\%
+ CPI-U Inflation Last Month (From Previous Year): `r round(LatestInflation,2)`\%

Lets get today's numbers in context with the past

## Unemployment

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

get_alfred_series("UNRATE") %>% 
  ggplot(aes(y =UNRATE, x = date)) + 
  geom_line() + ylab("Unemployment Rate") + xlab("")

```

## GDP Growth

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
get_alfred_series("A191RL1Q225SBEA") %>% 
  ggplot(aes(y =A191RL1Q225SBEA, x = date)) + 
  geom_line() + ylab("Annualized Growth Rate (Quarterly)") + xlab("")

```

## Inflation

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
get_alfred_series("CPALTT01USM659N") %>% 
  ggplot(aes(y =CPALTT01USM659N, x = date)) + 
  geom_line() + ylab("CPI-U Inflation (Annualized)") + xlab("")

```

## Where are we going with this?

Into the past

+ Talk about how we started with each of these measures
+ How our ideas have changed
+ Some controversies
+ Keep in mind that we rolled into the great depression with none of this, we made it up because we needed to.


## Unemployment

We came into the great depression with no national measure of unemployment.  We were guessing from many different sources.

+ While conducting the 1940 census
+ We tried a probability sample at the same time to see if the sample reflected the census.
+ That survey eventually became the **Current Population Survey** -- which is conducted monthly.
+ There is also an annual supplement with other questions.

## The Activity Concept

Finding if people are employed is easy -- We ask if people are **employeed for pay or profit**.

+ Assumes any job, part-time or full-time is good.
+ Assumes that the job is the best use of skills.


Unemployment is harder:

+ We could ask if people wanted a job, but that is more of a hypothetical that comes with no cost.
+ We what they **did** to find a job

$$Unemployment~Rate =\frac{Unemployed}{Unemployed + Employed}$$


## The First Survey

+ March 1940
+ Probability sampling worked.
+ Measured 15% unemployment, high, but lower than the estimated 25% the the peak of the Great Depression.

## 1942 Change in Concerns

+ Pearl Harbor
+ They were suddenly worried  about having enough workers
+ They added questions about who could work
    + Women
    + Children
+ Soon unemployment was about 1%.

## End of WW2

+ There was some worry that we would go back into the Great Depression
    + We didn't
    + Yeah, fiscal stimulus.
    
The worry and the idea that government spending can keep us out of depression was in many places, both technical writing and the arts.

+ Brave New World (1931)
+ Keynes's General Theory (1936)
+ 1984 (1949)

## The 50s

A few changes in the unemployment concepts:

+ Temporary layoff (< 30days) were now "unemployed"
+ Waiting to start a job (< 30 days) now "unemployed"

This helped make the measure more sensitive to output.

They also noticed that some people said that they were not looking for work because of the, “belief that no work was available in their line of work or community.”

## Seasonality

There are wide swings in employment, much Agriculture driven (1/5 work force then but 1/50th now.)

+ For much of the 40s they calculated unemployed/population -- a different rate than we report now.
+ They removed regular patterns so you can:     
    + compare unemployment month to month and not just
    + same month in previous year. 
    
## The 60s Begin

+ 1959, 2 years after the last recession, unemployment is still over 5%.
+ New recession in April 1960.
+ Accusations that methods magnified the growing unemployment problem
+ Seasonal adjustments were accused of being used for political purposes.

## Gordon Committee 1961

+ No manipulations.  Stats done well.
+ Added:
    + Looked in the last 4 weeks (specific)
    + Available to work in last 4 weeks.
    + Defined employed as 1 hr of pay or profit
+ Concerns about automation
    + Ask whey they were not looking.
    
## Gorden Committee (Con't)

Is unemployment a measure of hardship?

+ Indirect measure of hardship.
+ Not everyone who is looking, needs a job
    + Students
    + Housewives
+ Income data in supplemental survey asked income.
+ Poverty rates, based on threshold income and family size, were now also published as a hardship measures (1965).

## The 70s


+ The "significance" controversy and accusations of manipulation.
+ U-1 through U-7, the new measures of unemployment
+ Another commission

## Significance

+ February 1971 0.2% drop in unemployment rate
    + BLS "marginally significant"
    + Labor Sec. “of great significance.” 

+ March 1971
    + BLS “sort of mixed”
    + Labor Sec. "heartening"

+ July 1971 had a technical problem.

Accusations of politicization.


## This is Why We have Release Schedules

+ All dates are known a year ahead of time
+ Restricted access to pre-rease data
+ No public commentary till 1hr before

Trump violated this in [**June 2018**](https://www.wsj.com/articles/trump-says-he-is-looking-forward-to-jobs-report-1527855202)


## More measures

+ U-1 through U-4: Narrow definition of unemployed.
+ U-5: Official
+ U-6 through U-7: Broad

They all show similar trend and movement and have since been revised.


While similar issues have been in the [**news**](https://www.npr.org/2017/01/29/511493685/ahead-of-trumps-first-jobs-report-a-look-at-his-remarks-on-the-numbers) it is an old idea.


## The 90s

+ More focus on permanent and widespread job loss and definition of labor  force (Unemployed and Employed)
+ Discouraged worker measurement.


## The Current Definitions

+ U1: Percent of labor force unemployed 15 weeks or longer
+ U2: Percent of labor force who lost jobs or completed temporary work.
+ U3: Official Without jobs and have looked for work in past 4 weeks.
+ U4: U3 + Discouraged workers (Believe no work is available and stopped looking)
+ U5: U4 + Marginally attached (Stopped looking but think there may be jobs.)
+ U6: U5 + Part-time workers that want full-time but can't find it.

Currently:

+ [**Latest Release**](https://www.bls.gov/news.release/empsit.nr0.htm)
+ [**The Alternative Measures**](https://www.bls.gov/news.release/empsit.t15.htm)


## The CPI

+ It is older than our GDP measures
+ It is older than our unemployment measures
+ Goes back to 1913
+ It is embeded in law and is not revised when new data is available.


## The Fundimental Problem

How do you find average prices?

+ 

<!-- + Macroeconmoics sum more than parts.  Immergent properties -->
<!-- + Contrast great depression  and recession.  Answer, same track but we pulled out after about 15 months, something we didn't do for the depression -->
<!-- + What is a business cycle -->
<!-- + Trough to peak.   -->
<!-- + Who declares -->
<!-- + History of declarations, what did they look at ? -->

<!-- We use different models to think about long term growth and short term fluctuations -->

<!-- Inflation and deflation why do we get them -->
<!-- What are the different pains.  Why we like stability. -->
