---
title: "Ch 11: Income Expenditure Model"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Goals

+ Look at the simple income-expenditure model
+ Talk about some limitations
+ Examine one of the many multipliers

## A Simple Idea

+ People get disposable income, $yd$.
+ They spend some money just because, $a$.  This is called autonomous consumption.
+ They save some fraction, the marginal propensity to save ($MPS$), of each additional dollar of disposable income and spend other fraction, the marginal propensity to consume, ($MPC$), $MPC = 1 - MPS$$.

$$C = a + MPC * yd$$


## Picture

```{r Consumption Function, echo=FALSE, fig.cap='Increase Supply?', message=FALSE, warning=FALSE, paged.print=FALSE}

library(dplyr)
library(ggplot2)
library(Hmisc)

BaseConsumption <- ggplot(data.frame()) + 
  xlab("yd") +
  ylab("C") + 
  xlim(0,12) + ylim(0, 12) + theme_classic() 


BaseConsumption + geom_abline(slope = .6 , intercept = 3) +
  annotate("text", x = 9, y = 11, label = "Consumption Function", size = 2) 

```

## Of Note

+ MPC, the slope, does not tell you the average savings rate $\frac{yd - C}{yd}$$ but the fraction of each additional dollar of income that you save.
+ At low levels of income, you spend more than you save.
+ At high levels of income, you spend less than you save.
+ $a$ is the intercept and $MPC$ is the slope

## What Moves the Consumption Function?

+ Beliefs about the future
    + If you think your income will be higher in the future, spend more now and save less, i.e., shift up.
    + If you think your income will be lower in the future, spend less now and save more, i.e., shift down.
+ Wealth
    + Adequate or inadequate wealth depends on where you are in your earning years.
    + Booming stock market often indicates an increase in wealth, and an increase, shift up, in the expenditure function.
    
    
## Massive Simplifications

+ Why are you saving?
+ Why linear?
+ When do I die/retire/have my rainy day?

## Lets Build an Economy

+ C is as described, $C = a + MPC * yd$
+ No government or international sector
+ I is the same for all $yd$ but can move based on the interest rate.
    + It has two components, planned ($I_{planned}$) and unplanned ($I_{unplanned}$)
+ The aggregate expenditure function (AE) is the sum of C and $I_{planned}$.

$$AE = a + MPC * yd + I_{planned}$$

## A Principle Returns  

Principle #10: One person’s spending is another person’s income. 


+ We need to have expenditures, $AE$, equal to income, $yd$.
+ If $AG$ is higher than $yd$, then inventories shrink, $I_{unplanned}<0$
+ If $AG$ is lower than $yd$, then inventories expand, $I_{unplanned}>0$

## On the Graph

```{r Equilibrium, echo=FALSE, fig.cap='Increase Supply?', message=FALSE, warning=FALSE, paged.print=FALSE}

library(dplyr)
library(ggplot2)
library(Hmisc)



BaseConsumption + geom_abline(slope = .6 , intercept = 3) +
  annotate("text", x = 9, y = 11, label = "Consumption Function", size = 2) +
  geom_abline(slope = 1 , intercept = 0) +
  geom_vline(xintercept = 5) +
  geom_vline(xintercept = 10)

```


## Note

+ If $AE>yd$, i.e, when yd is low, inventories shrink
+ if $AE<yd$, i.e, when yd is high, inventories build
+ Inventory change is what makes the income expenditure equilibrium stable.
+ Often called the "Keynesian Cross"


## So, what is the equilibrium

$$
\begin{aligned}
yd = a + yd (MPC) + I_{planned}\\
(1-MPC)yd = a + I_{planned}\\
yd = \frac{a + I_{planned}}{(1-MPC)}
\end{aligned}
$$

## (One of the) Multiplier(s)

$$\frac{1}{1-MPC}$$

+ This is one of many 'multipliers' in introductory macroeconomics.
+ Often goes by, "Fiscal Policy Multiplier".  You will see why in later chapters.

## Numerical without Investment

$$
\begin{aligned}
yd = 3 + yd (.6)+ 0\\
(1-.6)yd = 3\\
yd = \frac{3}{(1-.6)} = 7.5
\end{aligned}
$$

## Here is where it gets odd

Lets start saving more because we think the economy is going to tank. $a$ drops from 3 to 2.

$$
\begin{aligned}
yd = 2 + yd (.6)+ 0\\
(1-.6)yd = 2\\
yd = \frac{2}{(1-.6)} = 5
\end{aligned}
$$

Equilibrium GDP dropped from 7.5 to 5.  This is called the paradox of thrift.

## The Paradox of Thrift

+ If you believe the economy will be worse, lower GDP, in the future
+ You act like it and save more now.
+ When you save more now, GDP falls.


## Next Up

Getting the Price Level and GDP to talk to each other.  The next two chapters are very graph heavy.

